---
title: "Shapefile to predictions"
author: "Benjamin Shenouda, Trinetta Chong, Eugenio Noda, Tamma Carleton"
date: '2022-06-24'
output: rmarkdown::github_document
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Making predictions from label data

This notebook demonstrates the steps in the process of making predictions starting from your own shapefile label data. The process can be divided into 7 steps

1.  Load shape file
2.  Create a grid according to the desired resolution at which you would like to generate predictions
3.  Create lat-lon crosswalk from the grid to the polygon, matching each point in the grid to corresponding polygon value
4.  Output the new grid label dataset to a CSV file for use in the MOSAIKS API File Query
5.  Using the lat-lon crosswalk from step 3, merge label data with the MOSAIKS features obtained through the API's File Query function
6.  Split merged dataset into train and test sets. In order to learn how visual information in the satellite imagery (as captured by the MOSAIKS features) relates to the label, perform a ridge regression on training dataset.
7.  Finally,  use the trained model to make predictions in the test set. Evaluate performance in the test set by comparing predictions to the label data.

## 1. Load shapefile

```{r}

suppressPackageStartupMessages(library(sf))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(mapview))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(rgdal))
suppressPackageStartupMessages(library(glmnet))
suppressPackageStartupMessages(library(patchwork))

```

Prepare directories and files

```{r}

input_dir <-  '/Users/trinettachong/Dropbox/Sherrie/mosaiks_api/mosaiks-demo/' #input path to working directory
label_shp_name <-  'ecu_soil_erosion_data/ecu_ica_landdegradation_geonode_jul2018.shp'
target_crs <-  4326 # epsg:4326, the projection of the MOSAIKS grid

label_shp <- st_read(paste0(input_dir, label_shp_name)) # read in shapefile using sf package
label_shp_reprojected <- st_transform(label_shp, crs = target_crs) # if needed, reproject shapefile to epsg:4326
```

Inspect shapefile and data

```{r}
plot(label_shp[,"ErosPr"], axes = T, main= "Soil Erosion") # Plot variable "soil erosion index" ErosPr
```

## 2. Create grid

In this step, we create a grid at a desired resolution, and constrain it by the bounding box of the label data. The specified resolution represents the desired resolution at which predictions are later made. 

A `target_resolution <- .01` signifies a 0.01 x 0.01 degree latitude-longitude grid, centered at 0.005 degree intervals (0.01 degrees is approximately equivalent to 1.11 km at the equator using an equal angle grid). This resolution is equivalent to that of the MOSAIKS API features. 

In many cases, one can make predictions at a finer resolution than the native resolution of input label data. For example, the native resolution of this label is at the canton administrative level (as seen in the shapefile plot). However, by overlaying our desired grid onto the shapefile and performing a spatial join, we are able to assign label values to each unique cell on a 0.01 x 0.01 degree grid. 

One should note that increasing the resolution will increase compute time. For illustrative purposes and to reduce compute time, this demo uses a slightly lower resolution of `target_resolution <- .05` (i.e. 0.05 x 0.05 degree grid, centered at 0.025 degree intervals)


```{r}

target_resolution <- .05 # Set target resolution. A resolution of .01 is equal to the .01 x .01 dense grid of the MOSAIKS API.

label_bbox <- st_bbox(label_shp_reprojected) # get bounding box of shapefile


# Creates grid of at specified resolution starting at the min x and min y coordinates of the label bounding box

grid <- st_make_grid(label_shp_reprojected, 
        cellsize = target_resolution, # cell size intervals for the grid, can be modified for higher or lower resolution -> .01 is equal to the MOSAIKS dense grid
        what = 'centers', # to render points at the centroids of each grid cell
        offset = c(round(label_bbox[1], 2), round(label_bbox[2]), 2)) # starting at the min x and y of the bbox rounded to 3 digit precision of the MOSAIKS grid

```

Plot grid over shapefile

```{r}

# Plot grid over shp with values
ggplot() +
  geom_sf(data = label_shp_reprojected, aes(fill = ErosPr), lwd = 0.09) +
  scale_fill_gradientn(colors = sf.colors()) + 
  geom_sf(data = grid, fill = 'transparent', lwd = 0.01) +
  coord_sf(datum = 4326)  +
  labs(x = "", y = "", fill = "Soil Erosion")
  

```

## 3. Spatially join the grid with label shapefile

In this step, we overlay the generated grid on the shapefile, and perform a spatial join in order to assign label values to each unique grid cell centroid.

```{r}

grid_sf <- st_as_sf(grid) # transform grid from sfc object to sf object for merging

label_df<- st_join(grid_sf, label_shp_reprojected[,"ErosPr"], join = st_within, left = FALSE) # spatial join grid over shapefile, "ErosPr" is the desired attribute column of the example dataset

```

Plot check

```{r, echo = FALSE, out.width="100%", out.height="100%"}
#plot(label_shp_reprojected[,"ErosPr"], axes = T) # plot shapefile
#plot(label_df[,"ErosPr"], axes = T) #plot newly generated grid data frame

shp_plot <- ggplot() +
  geom_sf(data = label_shp_reprojected, aes(fill = ErosPr)) +
  scale_fill_gradientn(colors = sf.colors()) +
  ggtitle("Shapefile") +
  labs(fill = "Soil Erosion")

grid_plot <- ggplot() +
  geom_sf(data = label_df, aes(color = ErosPr), size = 0.2) +
  scale_color_gradientn(colors = sf.colors()) +
  ggtitle("Gridded points") +
  labs(color = "Soil Erosion")

shp_plot + grid_plot + plot_layout(guides = "collect") & theme(legend.position = 'bottom')
```

## 4. Output the grid to CSV to obtain features from File Query

```{r, echo = FALSE}
# Split sfc point geometry into individual lon/lat values
label_df$lon <- st_coordinates(label_df)[,1] 
label_df$lat <- st_coordinates(label_df)[,2]

# Upon inspection, can confirm that no NAs are generated in label_df

st_write(label_df, paste0(input_dir, "ecu_erosion_label.csv")) # Write to csv - sfc point geomtery is automatically dropped. Results in 3 columns of label value, longitude, and latitude
```


# Query MOSAIKS Features from API

Using the newly-created .csv file of label data, MOSAIKS features can be obtained from the API's File Query function. Once features are downloaded, they can be merged with the label data.

## 5. Load and merge features and label data

Load and inspect downloaded features. 

Notice that each row represents the centroid of a unique grid cell, with the first two columns representing latitude and longitude coordinates, and subsequent columns representing K features (presently, K = 4000 features).

```{r}
#load features (this may take a few mins), and label data
features <- read.csv(paste0(input_dir,'Mosaiks_features.csv'), header = T, stringsAsFactors = F)
features <- features[with(features, order(Lat, Lon)),] #sort

head(features, c(5, 10)) # Limit output to 10 columns for sake of space. Actual dataframe contains 4002 columns.
```

Load and inspect label data. 

Similarly, each row in the label data represents the centroid of a unique grid cell, with the first column representing our label data (soil erosion), and the last two columns displaying latitude and longitude coordinates.

```{r}
label_filename = 'ecu_erosion_label.csv' # filename of the output csv from step 4
label <- read.csv(paste0(input_dir, label_filename), header = T, stringsAsFactors = F)
label$lat <- round(label$lat, 3) #3 decimal places
label$lon <- round(label$lon, 3) #3 decimal places
label <- label[with(label, order(lat, lon)),] #sort

head(label, 5)
```

Merge features spatially with label.

Here, we merge our downloaded features ("X") with our label ("y") based on the coordinates of the grid cell. This results in a single dataset with a vector of MOSAIKS features associated with each label observation. Each row in this dataset represents the label value and features for a given grid cell [**(resolution according to the specified grid resolution of the generated label grid)**]{.underline}

```{r}
#merge features with label
df <- left_join(label, features, by = c("lat"="Lat", "lon" = "Lon"))
print(paste(nrow(df), "observations in merged dataframe"))

#drop NAs
num_na <- nrow(df) - nrow(na.omit(df)) # get number of rows with NAs
df <- na.omit(df) # drop NAs from dataframe

print(paste0(num_na, " rows with NAs dropped. ", nrow(df), " observations in final dataframe."))

head(df, c(5, 10)) # limited to 10 columns for sake of space. Actual dataframe contains 4003 columns
```

## 6. Run a ridge regression of label on MOSAIKS features

This step learns how the visual information in the satellite imagery, as captured by the features, relates to the labels.

Before performing a ridge regression, we first split our data into train (80%) and test (20%). We will estimate the models on the train set and then evaluate predictions in the test set. This separation of train and test set is important to address issues related to overfitting.

```{r}
### Split data into training and test sets
set.seed(100) 

index <- sample(1:nrow(df), 0.8*nrow(df)) 

train <- df[index,] # Create the training data 
test <- df[-index,] # Create the test data

label_value <- "ErosPr" # set label attribute name in the dataframe, modify according to dataset

# Create X matricies and y vectors from the train and test sets
# Note that the glmnet function does not work with dataframes. Instead, they need to be converted to  numeric matrices  

X_train <- as.matrix(subset(train, select = -c(get(label_value))))
y_train <- train[,label_value]

X_test <- as.matrix(subset(test, select = -c(get(label_value))))
y_test <- test[,label_value]

#save test set lat/lons for plotting later
plotting_coords <- subset(test, select = c(lon,lat))
plotting_coords_train <- subset(train, select = c(lon,lat))
```

Within our training dataset, we choose the penalization parameter (i.e. lambda) in the ridge regression via k-fold cross-validation (here, k=5).

```{r}
# Train the ridge regression model, choosing the penalization parameter using cross-validation in the train set.
cv_ridge <- cv.glmnet(X_train, y_train, 
                      alpha = 0, # ridge regression
                      nfolds = 5)

# summarize chosen configuration 
optimal_lambda <- cv_ridge$lambda.min
print(paste0("optimal lambda: ", optimal_lambda))
```

## 7. Make predictions and evaluate performance

Next, we use the trained model to make predictions in the test set.

```{r}
#predict model
y_pred <- predict(cv_ridge, s = optimal_lambda, newx = X_test)
```

We can then evaluate performance in the test set by comparing predictions to the label data.

```{r}
# Compute R^2 from true and predicted values
SSE <- sum((y_pred - y_test)^2)
SST <- sum((y_test - mean(y_test))^2)
r2 <- (1 - SSE / SST)

print(paste0('r2: ', r2))
```

We can also make a scatter plot of labeled data against predictions.

```{r}
#clip bounds so min value = 0 because we know that label value cannot be below zero. 
y_pred[y_pred<0] <- 0

#combine lat/lons with test and predicted values into a dataframe
map_plot <- data.frame(plotting_coords, y_pred, y_test) 
names(map_plot) <- c('lon','lat','predicted','observed')

#scatterplot
ggplot(map_plot, aes(x=y_pred, y=y_test)) + 
geom_point(alpha = 0.5) +
geom_abline(intercept = 0, slope = 1, size = 0.8, alpha = 0.3) + # 45 degree line
labs(x = "Predicted", y = "Observed") 
```

Finally, we can compare the spatial distribution of label (i.e. observed) values with predicted values.

```{r, echo = FALSE, out.width="100%", out.height="100%"}
#make predictions in train set for plotting maps
y_pred_train <- predict(cv_ridge, s = optimal_lambda, newx = X_train)
y_pred_train[y_pred_train<0] <- 0
map_plot_train <- data.frame(plotting_coords_train, y_pred_train, y_train) 
names(map_plot_train) <- c('lon','lat','predicted','observed')
map_plot_train <- rbind(map_plot_train, map_plot)

scale_max <- max(map_plot_train$observed, map_plot_train$predicted)
#scale_max <- max(map_plot$observed, map_plot$predicted)

#plot observed 
observed <- ggplot() +
           geom_point(data = map_plot_train, 
                      aes(x=lon, y=lat, color= observed), 
                      size = 0.3, alpha = 0.8) +
           scale_color_distiller(palette = "YlGn", limits = c(0, scale_max)) +
           #labs(color = "observed soil erosion") +
           labs(x = "Longitude", y = "Latitude", title = "Observed", color = "Soil Erosion") +
           geom_sf(data =label_shp_reprojected, size = 0.5, alpha = 0.05)  

#plot predicted 
predicted <- ggplot() +
           geom_point(data = map_plot_train, 
                      aes(x=lon, y=lat, color= predicted), 
                      size = 0.3, alpha = 0.8) + 
           scale_color_distiller(palette = "YlGn", limits = c(0, scale_max)) +
           #labs(color = "predicted soil erosion") +
           labs(x = "Longitude", y = "Latitude", title = "Predicted", color = "Soil Erosion") +
           geom_sf(data =label_shp_reprojected, size = 0.5, alpha = 0.05)  
           

observed + predicted + plot_layout(guides = "collect") & theme(legend.position = 'bottom')

```
Example application: 
One could predict soil erosion risk at higher spatial resolution. For example, areas along the Napo River in east Ecuador face a higher risk of soil erosion.

```{r}
#load water areas in ecuador 
water_shp <- st_read(paste0(input_dir, 'ECU_wat/ECU_water_areas_dcw.shp')) # read in shapefile 
water_shp_reprojected <- st_transform(water_shp, crs = target_crs) # reproject water layer

#plot  
ggplot() +
geom_point(data = map_plot_train, 
  aes(x=lon, y=lat, color= predicted), 
  size = 3, alpha = 0.9) + 
scale_color_distiller(palette = "YlGn", limits = c(0, scale_max)) +
labs(x = "Longitude", y = "Latitude", title = "Napo River Basin", color = "Soil Erosion") +
geom_sf(data =label_shp_reprojected, size = 0.3, alpha = 0.01, color = "grey39") +
geom_sf(data =water_shp_reprojected, size = 0.1, alpha = 0.1, color = "blue1") + #water
  coord_sf(
  xlim = c(-78, -75),
  ylim = c(-1.2, 0),
)
```

